---
title: "Project 1: Activity Monitoring Devices"
author: "Roberto Samaniego Ferrabone"
date: "Friday, June 14, 2015"
---

This is Reproducible Research's Project 1 and makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The purpose of this Project is to use R Markdown with Knitr and R code chunks, to generate fully reproducible documents.

#### Loading and preprocessing the data.

Download activity.csv file from the Course Project web site described, which include the next columns or variables:  
- Steps: Number of steps taking in a 5-minute interval (missing values are coded as NA).  
- Date: The date on which the measurement was taken in YYYY-MM-DD format.  
- Interval: Identifier for the 5-minute interval in which measurement was taken.  


```{r Reading Activity Monitoring Devices file, echo=TRUE}
activity_ds <- read.csv("activity.csv", header = TRUE, sep = ",")
```

#### Mean total number of steps taken per day.
Ignore missing values in this part.
- Calculate the total number of steps taken per day.  
-If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day.  
- Calculate and report the mean and median of the total number of steps taken per day.  

```{r Steps taken per day Histogram, Mean and Median, echo=TRUE}
activity_subset <- na.omit(activity_ds)

steps_day_ds <- aggregate(steps ~ date, activity_subset, sum)
hist(steps_day_ds$steps, main = paste("Total Steps taken per Day"), col="lightblue", xlab="Number of Steps", ylab="Frequency")

varmean <- mean(steps_day_ds$steps)
varmedian <- median(steps_day_ds$steps)
```

Total mean of steps taken per day: `r varmean`.  
Total median of steps taken per day: `r varmedian`.

#### Average daily activity pattern.
- Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).  
- Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  

```{r Average Daily Activity Pattern, echo=TRUE}
average_daily_ds <- aggregate(steps ~ interval, activity_subset, mean)

plot(average_daily_ds$interval,average_daily_ds$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Daily Activity Pattern")
```

#### Imputing missing values.
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).  
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.  
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?  

```{r Imputing missing values, echo=TRUE}
varmissing <- sum(is.na(activity_ds$steps))
```

Missing values: `r varmissing`.

```{r Histogram for Total Number of Steps with Missing Values, echo=TRUE}
imputed_ds <- activity_ds
is_na <- is.na(imputed_ds$steps)
average_ds <- tapply(activity_subset$steps, activity_subset$interval, mean)
imputed_ds$steps[is_na] <- average_ds[as.character(imputed_ds$interval[is_na])]

daily_ds <- tapply(imputed_ds$steps, imputed_ds$date, sum, na.rm=TRUE)

hist(x=daily_ds,
     col="lightblue",
     breaks=20,
     xlab="Number of steps",
     ylab="Frequency",
     main="Daily total of Steps Taken per Day including Missing Values")

varimpmean <- mean(daily_ds)
varimpmedian <- mean(daily_ds)
```

Missing values: `r varimpmean`.  
Missing values: `r varimpmedian`.  

#### Differences in activity patterns between weekdays and weekends.

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.  
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.  

```{r Differences in Activity Pattern between weekdays and weekends, echo=TRUE}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
imputed_ds$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_ds$date)),weekdays), "Weekday", "Weekend"))

diff_ds <- aggregate(steps ~ interval + dow, imputed_ds, mean)

library(lattice)

xyplot(diff_ds$steps ~ diff_ds$interval | diff_ds$dow, main="Differences in Activity Pattern between weekdays and weekends",xlab="Interval", ylab="Steps Taken per Day",layout=c(1,2), type="l")
```
